---
title: "Exercises"
editor: visual
---

**Instructions**
As you complete the exercises keep your cursor in the exercise code chunk and ensure to run all chunks above the exercise chunk (PC: ctrl + alt + shift + P or Mac: cmd + alt + shift + P) followed by the current code chunk you are on (PC: ctrl + shift + enter, or Mac: cmd + shift + return), or _render_ the whole document (PC: ctrl + shift + K, or Mac: cmd + shift + K).

## Exercise 1

Create a box plot using `complete_old` of plot_type vs weight with the following features:

- Colour inside the box based on plot_type
- Boxes are 50% transparent
- Box outlines and outlier points are `grey75`
- Set the theme to be black and white
- Angle the x-axis text 45 degrees and hjust so the end of the label is inline with the tick
- Give nice labels to the axes and a title

Fix the FIXMEs and then switch the top line of the code block to `#| eval: true`. After which execute the code block (PC: ctrl + shift + enter, or Mac: cmd + shift + return)

```{r}
#| eval: false

library(ggplot2)
library(ratdat)

ggplot(data = FIXME, mapping = aes(x = FIXME, 
                                   y = FIXME, 
                                   fill = FIXME)) + 
  geom_boxplot(FIXME)+
  theme_FIXME() +
  theme(FIXME) +
  labs(FIXME)
```


## Exercise 2



Using the `surveys` dataset:

- remove `NA`'s from the `hindfoot_length`
- remove rows missing `sex`
- find the minimum, mean, and maximum `hindfoot_length`, and number of observations for each `species` 
- order the results by the longest `hindfoot_length`


Fix the FIXMEs and then switch the top line of the code block to `#| eval: true`. After which execute the code block (PC: ctrl + shift + enter, or Mac: cmd + shift + return)

```{r}
#| eval: false

library(tidyverse)
surveys <- read_csv("data/surveys_complete_77_89.csv")


FIXME %>% 
  filter(FIXME) %>% 
  group_by(FIXME) %>% 
  summarise(FIXME) %>% 
  FIXME(desc(FIXME))
```


