---
title: "Introduction to R"
warning: false
toc: true
---



```{r}
#| include: false

options(max.print = 100)
```


## Set up

### Install R & RStudio

On personal or unrestricted machines, install [R](https://cran.r-project.org/bin/windows/base/) and then [RStudio](https://posit.co/download/rstudio-desktop/) directly.

On restricted/organisational machines (e.g. your University or workplace), install both R and RStudio (in that order) from your institutional software repository.

Open RStudio, navigate to the top Tools menu > Install Packages, then enter `tidyverse,ratdat,glue` and click install. It may take a minute or two to install, and you’ll see lots of text printed in the console. Once finished, you will see a blue ‘>’ symbol in the bottom left console pane.

## 1. R & RStudio

### Programming vs point-and-click

R is a programming language as well as software that runs R code.

RStudio is a popular software interface that can make it easier to write R scripts and interact with R. 

Programming languages have many benefits over point and click software:

- Analysing your data with a series of written commands provides a record of exactly what you've done, aiding transparency and research integrity.
- When these commands are saved into a script file, they can easily be rerun on similar data, saving you valuable time and opening up new possibilities.
- R produces high-quality graphics, has a package for almost any type of analysis you'll want to perform, and has a large and welcoming user community for help and guidance.

### Navigating RStudio

RStudio has 4 panes, clockwise from top left:

- Source: Displays scripts, data, and other kinds of files.
- Environment: Displays the objects in your current R session.
- Files/Plots/Help/Viewer: Displays files on your computer, plots, and help documentation.
- Console: Where you interact directly with R.

### Projects

It's best to keep related files and analyses together in project folders. You can then set up an RStudio project you're working on, which provides some useful features when working with R.

Portable project stuff

### Console vs Script

You can run commands directly in the R console, or you can write them into an R script.

Console:

- Where code is run/executed
- Type commands at the prompt (`>` symbol) 
- Press <kbd>Enter</kbd> to execute the command and print the result
- Can't access what you did after closing RStudio

Script:

- A list of R commands in a plain text file with a `.R` extension
- `File → New File → R Script` or `+` button in the top left corner of RStudio
- `Cmd+Enter` (Mac) or `Ctrl+Enter` (Windows) will run the line of code that your cursor is on, or which is highlighted 
- You can leave comments with `#` 
- Commands are saved and can be rerun later

## 2. Plotting with ggplot2

Load the packages we need with `library()`. We'll talk more about functions later.

```{r message=FALSE}
library(ggplot2)
library(ratdat)
```

**`ggplot`** is a popular plotting package. Plots made with this packages are built step by step by adding new layers.

The `ratdat` package contains data from the [Portal Project](https://github.com/weecology/PortalData), a long-term dataset from Portal, Arizona, in the Chihuahuan desert. 

We will be using a dataset called `complete_old`.

```{r message=FALSE}
?complete_old
```

To build a plot we can start with a basic template:

`ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()`

We need to specify everything within `<>`.

```{r bare-plot}
ggplot(data = complete_old)
```

The plot is blank because we haven't told `ggplot()` which variables in the data we want to plot. 

```{r plot-with-axes}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length))
```

We haven't specified how we want the data to be displayed. We do this using `geom_` functions, which specify the type of `geom`etry we want, such as points, lines, or bars. We can add a `geom_point()` layer to our plot by using the `+` sign.

```{r scatter-plot}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

When we have overlapping points we can adjust the transparency of the points using the `alpha` argument, which takes a value between 0 and 1:

```{r change-alpha, warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2)
```

Change the color of points:

```{r change-color, warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2, color = "blue")
```

Change the colour of points to take on the value of another variable:

```{r color-plot-type, warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) +
  geom_point(alpha = 0.2)
```

Make a boxplot with `plot_type` on the x axis:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot()
```

Change the color of plots using `fill` and remove legend:

```{r warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot(show.legend = FALSE)
```

Change theme by specifying a built-in theme called `theme_bw()`:

```{r warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot(show.legend = FALSE) +
  theme_bw()
```

Give the plot a title and change the x and y labels:

```{r warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot(show.legend = FALSE) +
  theme_bw() +
  labs(title = "Rodent Size By Plot Type",
       x = "Plot Type",
       y = "Hindfoot Length (mm)")
```

Centre the plot title:

```{r warning=FALSE}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot(show.legend = FALSE) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Rodent Size By Plot Type",
       x = "Plot Type",
       y = "Hindfoot Length (mm)")
```

Save the plot by assigning it to an object called `final_plot` and using `ggsave()`:

```{r warning=FALSE}
final_plot <- ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot(show.legend = FALSE) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Rodent Size By Plot Type",
       x = "Plot Type",
       y = "Hindfoot Length (mm)")

ggsave(plot = final_plot, filename = "plots/rodent-size-plot-type.png",
       width = 8, height = 6)
```

By default, ggplot dimensions are specified in inches, and images are 300 DPI. So this image will have a resolution of 2400 pixels wide by 1800 pixels high.

:::::: {.callout-caution title="Exercise (5 min)"}
Create a box plot using `complete_old` of plot_type vs weight with the following features:

- Colour inside the box based on plot_type
- Boxes are 50% transparent
- Box outlines and outlier points are `grey75`
- Set the theme to be black and white
- Angle the x-axis text 45 degrees and hjust so the end of the label is inline with the tick
- Give nice labels to the axes and a title

```{r}
#| eval: false

ggplot(data = FIXME, mapping = aes(x = FIXME, 
                                   y = FIXME, 
                                   fill = FIXME)) + 
  geom_boxplot(FIXME)+
  theme_FIXME() +
  theme(FIXME) +
  labs(FIXME)

```

::: {.callout-note collapse="true" title="Example solution"}
```{r}
# create a plot
ggplot(data = complete_old, mapping = aes(x = plot_type, 
                                          y = weight, 
                                          fill = plot_type)) + 
  geom_boxplot(alpha = 0.5, 
               colour = 'grey75')+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  labs(title = "Rodent Weight by Plot Type",
       x = "Plot Type",
       y = "Weight (g)")

```
:::
::::::

## 3. Programming basics

### Vectors

There are 4 main types of vectors:

1. `"character"` for text. Each entry is wrapped in quotes. In other programming languages, this type of data may be referred to as "strings".

2. `"integer"` for integers. All the numeric values in `complete_old` are integers. You may sometimes see integers represented like `2L` or `20L`. The `L` indicates to R that it is an integer, instead of the next data type, `"numeric"`.

3. `"numeric"`, aka `"double"`, vectors can contain numbers including decimals. Other languages may refer to these as "float" or "floating point" numbers.

4. `"logical"` for `TRUE` and `FALSE`, which can also be represented as `T` and `F`. In other contexts, these may be referred to as "Boolean" data. Note they are not wrapped in quotes.

Vectors can only be of a **single type**. 

To create a vector from scratch, we can use the `c()` function, putting values inside, separated by commas. Vectors are the basic building blocks of all data in R.

### Data.frames

Data.frames are made up of vectors; each column in a data.frame is a vector. 

We can see more information about `complete_old` by using the structure function:

```{r}
str(complete_old)
```

The `$` in front of each variable is an operator that allows us to select individual columns from a data.frame.


```{r dollar-subsetting}
complete_old$year
```


We get back all values in the `year` column.

### Objects and assignment

We can assign a value to an object by naming the object and using the assignment operator (`<-`):

```{r }
x <- 5
```

We can also assign objects to other objects:

```{r }
y <- x
```

If we now overwrite the value of `x`, the value of `y` will not change:

```{r }
x <- 10
y
```

### Functions and arguments

Functions take arguments, and some arguments are required. Optional arguments are called options. 

The `round()` function rounds a number. Lets take a look at how it works:

```{r }
?round()
```

It requires a number to round (`x`), and optionally the number of digits to round to (`digits`):

```{r }
round(x = 3.14159, digits = 1)
```
If we don't provide `digits` it will default to 0:

```{r }
round(x = 3.14159)
```

If we provide arguments in the order expected we don't have to name them (but it's good to when starting out):

```{r }
round(3.14159, 1)
```

### Reading in data

Up until now, we've been working with the `complete_old` dataframe contained in the `ratdat` package. However, you'll typically want to access data stored somewhere on your computer as files. Our project contains a data file in the `/data` directory, and we're going to read it in now. 

First, let's load the Tidyverse package. This contains many other packages, including some we've used already like ggplot2, but it's usually more convenient to load all of them at once.

```{r load-tidyverse-package, message=FALSE}
library(tidyverse)
```

**`tidyverse` vs. base R**

In R, there are often many ways to get a job done. The phrase **base R** is used to refer to approaches that utilize functions contained in R's default packages. There are some key base R approaches we will not be teaching. These include square bracket subsetting and base plotting. You may come across code written by other people that looks like `surveys[1:10, 2]` or `plot(surveys$weight, surveys$hindfoot_length)`, which are base R commands. 

The `tidyverse` packages share a similar syntax and philosophy, making them consistent and producing highly readable code. They are also very flexible and powerful, and tend to have very clear documentation written with novice users in mind. 

Because we're using an RStudio project, we can specify the relative file path when reading in our data (relative to the project folder):

```{r read-csv}
surveys <- read_csv("data/surveys_complete_77_89.csv")
```

### Subsetting 

Two of the most commonly used functions for manipulating data are `select()`, which selects certain columns of a data.frame, and `filter()`, which filters out rows according to certain criteria.

The first argument for `select()` is the name of the data.frame, and the rest of the arguments are *unquoted* names of the columns you want:

```{r select}
select(surveys, plot_id, species_id, hindfoot_length)
```

The columns are arranged in the order we specified inside `select()`.

Put a `-` in front of the column you want to exclude:

```{r select-minus}
select(surveys, -record_id, -year)
```

To select the 3rd, 4th, 5th, and 10th columns, we could run the following code:

```{r select-vector}
select(surveys, c(3:5, 10))
```

The `filter()` function is used to select rows that meet certain criteria. To get all the rows where the value of `year` is equal to 1985:

```{r filter}
filter(surveys, year == 1985)
```

The `==` sign means "is equal to". There are several other operators we can use: >, >=, <, <=, and != (not equal to). Another useful operator is `%in%`, which asks if the value on the lefthand side is found anywhere in the vector on the righthand side. For example, to get rows with specific `species_id` values, we could run:

```{r filter-in}
filter(surveys, species_id %in% c("RM", "DO"))
```

We can also use multiple conditions:

```{r filter-multiple}
filter(surveys, year <= 1988 & !is.na(hindfoot_length))
```

We get rows with a year less than or equal to 1988 and whose hindfoot length values are not `NA`. The `!` before the `is.na()` function means "not".

### Piping

What happens if we want to both `select()` and `filter()` our data? 

An elegant solution to this problem is an operator called the **pipe** `%>%`. You can insert it by using the keyboard shortcut `Shift+Cmd+M` (Mac) or `Shift+Ctrl+M` (Windows). 

```{r filter-select-pipe}
surveys %>% 
  select(-day) %>% 
  filter(month >= 7)
```

The pipe takes the object on the lefthand side and inserts it as the first argument of the function on the righthand side. By putting each of our functions onto a new line, we can build a nice, readable **pipeline**. 

We can assign this final product to an object:

```{r pipeline-assignment}
surveys_sub <- surveys %>% 
  select(-day) %>% 
  filter(month >= 7)
```

Build a pipeline step by step prior to assignment. Add functions to the pipeline as you go, with the results printing in the console for you to view. Once you're satisfied with your final result, go back and add the assignment arrow statement at the start. This approach is very interactive, allowing you to see the results of each step as you build the pipeline, and produces nicely readable code.

Another common task is creating a new column based on values in existing columns. For example, to add a new column that has the weight in kilograms instead of grams:

```{r mutate}
surveys %>% 
  mutate(weight_kg = weight / 1000)
```

You can create multiple columns in one `mutate()` call, and they will get created in the order you write them. This means you can even reference the first new column in the second new column:

```{r mutate-multiple}
surveys %>% 
  mutate(weight_kg = weight / 1000,
         weight_lbs = weight_kg * 2.2)
```

### Grouping

Many data analysis tasks can be achieved using the split-apply-combine approach: you split the data into groups, apply some analysis to each group, and combine the results in some way. 

`group_by()` takes a data.frame and the name of one or more columns with categorical values that define the groups. `summarize()` then collapses each group into a one-row summary of the group, giving you back a data.frame with one row per group. The syntax for `summarize()` is similar to `mutate()`, where you define new columns based on values of other columns. 

To calculate the mean weight of all our animals by sex:

```{r group-by-summarize}
surveys %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = T))
```

We can define multiple columns in one `summarize()` call. The function `n()` will count the number of rows in each group.

```{r summarize-multiple}
surveys %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = T),
            n = n())
```

You will often want to create groups based on multiple columns. For example, we might be interested in the mean weight of every species + sex combination.

```{r group-by-multiple}
surveys %>% 
  group_by(species_id, sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = T),
            n = n())
```

Our resulting data.frame is much larger, since we have a greater number of groups. We also see a strange value showing up in our `mean_weight` column: `NaN`. This stands for "Not a Number", and it often results from trying to do an operation a vector with zero entries. If a particular group (like the AB species ID + `NA` sex group) has **only** `NA` values for weight, then the `na.rm = T` argument in `mean()` will remove **all** the values prior to calculating the mean. The result will be a value of `NaN`. Since we are not particularly interested in these values, let's add a step to our pipeline to remove rows where weight is `NA` **before** doing any other steps. This means that any groups with only `NA` values will disappear from our data.frame before we formally create the groups with `group_by()`.

```{r filter-group-by}
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id, sex) %>% 
  summarize(mean_weight = mean(weight),
            n = n())
```

It's often useful to take a look at the results in some order, like the lowest mean weight to highest. We can use the `arrange()` function for that:

```{r arrange}
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id, sex) %>% 
  summarize(mean_weight = mean(weight),
            n = n()) %>% 
  arrange(mean_weight)
```

If we want to reverse the order, we can wrap the column name in `desc()`:

```{r arrange-desc}
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id, sex) %>% 
  summarize(mean_weight = mean(weight),
            n = n()) %>% 
  arrange(desc(mean_weight))
```


:::::: {.callout-caution title="Exercise (5 min)"}

Using the `surveys` dataset:

- remove `NA`'s from the `hindfoot_length`
- remove rows missing `sex`
- find the minimum, mean, and maximum `hindfoot_length`, and number of observations for each `species` 
- order the results by the longest `hindfoot_length`


```{r}
#| eval: false

FIXME %>% 
  filter(FIXME) %>% 
  group_by(FIXME) %>% 
  summarise(FIXME) %>% 
  FIXME(desc(FIXME))
```

::: {.callout-note collapse="true" title="Example solution"}
```{r}

surveys |>
  filter(!is.na(hindfoot_length), !is.na(sex)) %>% 
  group_by(species) %>% 
  summarise(
    n = n(), 
    min_hindfoot_length = min(hindfoot_length),
    mean_hindfoot_length = mean(hindfoot_length),
    max_hindfoot_length = max(hindfoot_length)
  ) %>% 
  arrange(desc(max_hindfoot_length))

```
:::
::::::


## 4. The power of functions

### Problem: Creating multiple plots

Let's say you need to create separate boxplots showing the variation in hindfoot length based on plot type, but for each species. One way to approach this problem would be to filter to a single species, and pipe the filtered data into a ggplot call. For exmaple, to do this for the rodent species marked "NL":

```{r}
complete_old %>%
  filter(species_id == "NL" & !is.na(hindfoot_length)) %>%
  ggplot(aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  theme_bw()
```

But repeating this chunk of code many times to produce plots for all species is inefficient, takes up a lot of room, and makes it hard to tweak aspects of the plots once you've written the code.

### Writing a function

When working with programming languages, writing your own functions allows you automate sets of operations together in a much more efficient way.

To write a function you need to first analyse your code to figure out what parts are constant and what parts vary. Let's look at the previous bit of code and replace the parts that could change with ___:

```{r eval=FALSE}
complete_old %>%
  filter(species_id == "___" & !is.na(hindfoot_length)) %>%
  ggplot(aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  theme_bw()
```

This bit of code requires a species ID code to create a plot based on that species, but the other parts remain constant.

When writing a function, start with this basic template:

```{r eval=FALSE}
name <- function(arguments) {
  body
}
```

Let's call our example function plot_species, let's name our argument sp_id to refer to species_id, and let's replace the body with the appropriate code from above:

```{r}
plot_species <- function(sp_id) {
  complete_old %>%
    filter(species_id == sp_id & !is.na(hindfoot_length)) %>%
    ggplot(aes(x = plot_type, y = hindfoot_length)) +
    geom_boxplot() +
    theme_bw()
}
```

Lets try it out for a single species ID:

```{r}
plot_species("RM")
```

Now that we know it works, let's use it to create plots for all 36 species. We're going to use the `map()` function from the {purrr} package (included in {tidyverse}) to do this. This function takes a list or vector (in our case unique values from the species_id column), and a function to apply to all values (in our case, the plot function we just created):

```{r output=FALSE}
map(unique(complete_old$species_id), plot_species)
```

Let's modify our function to give each plot a title to label which species_id is being plotted. To do this we're going to use a package called glue, so we'll need to load that too:

```{r output=FALSE}
library(glue)

plot_species <- function(sp_id) {
  complete_old %>%
    filter(species_id == sp_id & !is.na(hindfoot_length)) %>%
    ggplot(aes(x = plot_type, y = hindfoot_length)) +
    geom_boxplot() +
    theme_bw() +
    labs(title = glue("Variation in hindfoot length based on plot type for {sp_id} species"))
}

map(unique(complete_old$species_id), plot_species)
```

We can stop R from creating plots for species which only contain `NA` values by filtering valid species_id codes. First filter all rows out if they contain an `NA` value for hindfoot_length, then use `pull()` to grab the remaining species_id codes, and use `unique()` to end up with a vector of the unique species_id codes that have hindfoot_length measurements:

```{r}
valid_species <- complete_old %>%
  filter(!is.na(hindfoot_length)) |> 
  pull(species_id) |> 
  unique()
```

Now we can perform the map again on valid species_id codes:

```{r output=FALSE}
map(valid_species, plot_species)
```

Let's modify our function one more time to save each plot as a separate file with a filename based on the species code within the plot:

```{r output=FALSE}
plot_species <- function(sp_id) {
  complete_old %>%
    filter(species_id == sp_id & !is.na(hindfoot_length)) %>%
    ggplot(aes(x = plot_type, y = hindfoot_length)) +
    geom_boxplot() +
    theme_bw() +
    labs(title = glue("Variation in hindfoot length based on plot type for {sp_id} species"))
  
  ggsave(filename = glue("plots/hflength-box-{sp_id}.png"), plot = last_plot())
}

map(valid_species, plot_species)
```

## Resources

[Data Analysis and Visualization in R for Ecologists](https://datacarpentry.github.io/R-ecology-lesson/)

[R for Data Science](https://r4ds.hadley.nz/)

[Tidyverse documentation](https://www.tidyverse.org/packages/)

[Posit Forums](https://forum.posit.co/)

[Stack Exchange](https://stats.stackexchange.com/questions/tagged/r)